---
title: "Figures for FINESST"
author: "Maddie Berger"
date: "2025-01-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(lubridate)

cwb_raw <- read_csv(file.path(here("data/advisory-export.csv"))) %>%
  clean_names()



brown_water_clean <- cwb_raw %>% 
  mutate(date = format(mdy_hms(issuance_date), "%m/%d/%Y")) %>% 
  mutate(cause = str_to_lower(cause)) %>% 
  mutate(cause_grouped = case_when(
    str_detect(cause,"rain") == T ~ "rain",
    str_detect(cause,"storm") == T ~ "storm",
    str_detect(cause,"weather") == T ~ "storm",
    str_detect(cause, "river") == T ~ "river or stream breach",
    str_detect(cause, "stream") == T ~ "river or stream breach",
    str_detect(cause, "steam") == T ~ "river or stream breach",
    str_detect(cause, "flood") == T ~ "river or stream breach",
    str_detect(cause, "wave") == T ~ "high surf or wave action",
    str_detect(cause, "surf") == T ~ "high surf or wave action",
    str_detect(cause, "erosion") == T ~ "erosion",
    str_detect(cause, "landslide") == T ~ "erosion",
    str_detect(cause, "construction") == T ~ "construction",
    str_detect(cause, "hurricane") == T ~ "storm",
    str_detect(cause,"resuspension") == T ~ "resuspension",
    str_detect(cause, "runoff") == T ~ "runoff",
    str_detect(cause, "debris") == T ~ "debris",
    TRUE ~ "unknown"
  )) %>% 
  rowid_to_column()
  

```


# Which causes overall have been reported most often?

```{r}

causes_frequency <- brown_water_clean %>% 
  group_by(cause_grouped, county) %>% 
  summarize(
    num_alerts = length(rowid)
  )

# bar chart by cause and county

cause_pal <- c(
  "unknown" = "azure2",
  #"2012" = "azure3",
  #"2013" = "paleturquoise",
  "high surf or wave action"= "paleturquoise1",
  #"2015" = "paleturquoise3",
  "river or stream breach" = "cadetblue",
  "rain" = "skyblue1",
  "resuspension"= "skyblue3",
  "storm" = "skyblue4",
  "debris" = "lightgoldenrod1",
  "runoff" = "gold1",
 # "2022" = "goldenrod1",
  "erosion" = "darkgoldenrod2",
  "construction" = "coral")

cause_bar <- causes_frequency %>% 
  ggplot(aes(x = county, y = num_alerts, fill = cause_grouped))+
  geom_bar(stat = "identity", position = "stack")+
  scale_y_continuous(expand = c(0,0))+
  theme_bw()+
  labs(x = "County",
       y = "Number of Brown Water Alerts",
       fill = "Reported cause",
       title = "Brown water alerts 2010 - 2024")+
  scale_fill_manual(values = cause_pal,
                    limits = names(cause_pal))


cause_bar

ggsave(
  filename = "bwa_bar.png",
  path = here("figs"),
  dpi = 300
)


```


# Timeseries data

Let's look the time component in two ways 
- for seasonal patterns and to look at differences between years, graph month on the x axis, alerts on y, with individual lines representing years
- for changes in cause over time, we'll graph year on the x axis, alerts on y, and individual lines representing causes
- we will repeat for month


```{r}

# create new month column, a special column to format the x axis, and create a factored year column

bwa_time <- brown_water_clean %>%
  group_by(date) %>% 
  summarise(
    count = length(rowid)
  ) %>% 
  mutate(date = as.Date(mdy(date, tz = "HST"))) %>% # key for graphing
  mutate(year = factor(year(date))) %>% 
  mutate(month = month(date, label = TRUE, abbr = TRUE)) %>%
  #mutate(month_name = format(as.Date(paste0(month, "-01"), "%m-%d"), "%B")) %>% # update 1/25 - not working
  mutate(date_axis = update(date, year = 1)) %>% 
  mutate(count = as.numeric(count))

bwa_month <- bwa_time %>% 
  group_by(month,year) %>% 
  summarize(
    count = sum(count)
  )

year_pal <- c(
  "2010" = "azure1",
  "2011" = "azure2",
  "2012" = "azure3",
  "2013" = "paleturquoise",
  "2014"= "paleturquoise1",
  "2015" = "paleturquoise3",
  "2016" = "cadetblue",
  "2017" = "skyblue1",
  "2018"= "skyblue3",
  "2019" = "skyblue4",
  "2020" = "lightgoldenrod1",
  "2021" = "gold1",
  "2022" = "goldenrod1",
  "2023" = "darkgoldenrod2",
  "2024" = "coral")

bwa_alerts_pyear <- bwa_month %>% 
  ggplot(data =., mapping = aes(
    x = month,
    y = count,
    color = year,
    group = year,
  ))+
  geom_point(aes(color = year))+
  geom_line(aes(color = year))+
  #geom_bar(stat = "identity", position = "stack")+
  #scale_x_date(date_labels = "%b", date_breaks = "1 month")+
  scale_color_manual(values = year_pal,
                     limits = names(year_pal))+
  theme_minimal()
  


bwa_alerts_pyear

```

Have causes shifted over the years? Do they change per months?

```{r}

# lets look at alerts per year by cause and by month by cause

bwa_year_cause <- brown_water_clean %>%
  mutate(date = as.Date(mdy(date, tz = "HST"))) %>% # key for graphing
  mutate(year = factor(year(date))) %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE)) %>% 
  group_by(cause_grouped,year) %>% 
  summarize(
    count = length(unique(rowid))
  )

  
  
bwa_alerts_pyear_cause <- bwa_year_cause %>% 
  ggplot(data =., mapping = aes(
    x = year,
    y = count,
    fill = cause_grouped,
    group = cause_grouped,
  ))+
  geom_bar(stat = "identity", position = "stack")+
  #geom_bar(stat = "identity", position = "stack")+
  #scale_x_date(date_labels = "%b", date_breaks = "1 month")+
  scale_fill_manual(values = cause_pal,
                     limits = names(cause_pal))+
  theme_minimal()
  


bwa_alerts_pyear_cause

# by month?

bwa_month_cause <- brown_water_clean %>%
  mutate(date = as.Date(mdy(date, tz = "HST"))) %>% # key for graphing
  mutate(year = factor(year(date))) %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE)) %>% 
  group_by(cause_grouped,month) %>% 
  summarize(
    count = length(unique(rowid))
  )

bwa_alerts_pmonth_cause <- bwa_month_cause %>% 
  ggplot(data =., mapping = aes(
    x = month,
    y = count,
    fill = cause_grouped,
    group = cause_grouped,
  ))+
  geom_bar(stat = "identity", position = "stack")+
  #geom_bar(stat = "identity", position = "stack")+
  #scale_x_date(date_labels = "%b", date_breaks = "1 month")+
  scale_fill_manual(values = cause_pal,
                     limits = names(cause_pal))+
  theme_minimal()
  


bwa_alerts_pmonth_cause

# last one - year by island

county_pal <- c(
  #"unknown" = "azure2",
  #"2012" = "azure3",
  #"2013" = "paleturquoise",
  #"high surf or wave action"= "paleturquoise1",
  "Kauai" = "paleturquoise3",
  "Hawaii" = "cadetblue",
  #"Hawaii" = "skyblue1",
  #"resuspension"= "skyblue3",
  #"storm" = "skyblue4",
  "Honolulu" = "lightgoldenrod1",
  #"runoff" = "gold1",
 # "2022" = "goldenrod1",
  "Maui" = "darkgoldenrod2"
  #"Hawaii" = "coral"
 )
  


bwa_year_island <- brown_water_clean %>%
  mutate(date = as.Date(mdy(date, tz = "HST"))) %>% # key for graphing
  mutate(year = factor(year(date))) %>%
  mutate(month = month(date, label = TRUE, abbr = TRUE)) %>% 
  group_by(county,year) %>% 
  summarize(
    count = length(unique(rowid))
  )

bwa_alerts_pcounty_year <- bwa_year_island %>% 
  ggplot(data =., mapping = aes(
    x = year,
    y = count,
    fill = county,
    group = county
  ))+
  geom_bar(stat = "identity", position = "stack")+
  #geom_bar(stat = "identity", position = "stack")+
  #scale_x_date(date_labels = "%b", date_breaks = "1 month")+
  scale_fill_manual(values = county_pal,
                     limits = names(county_pal))+
  theme_minimal()+
  labs(x = "Year",
       y = "Number of Brown Water Alerts",
       fill = "County",
       title = "Brown water alerts issued for the MHI 2010 - 2024")


bwa_alerts_pcounty_year


ggsave(
  bwa_alerts_pcounty_year,
  filename = "bwa_pyear_pcounty.png",
  path = here("figs"),
  dpi = 300
)

```


